---
title: "_Gene Expression Signatures in Blood that Predict Mortality in a West African Sepsis Cohort Define Host Response Phenotypes_"
output:
  pdf_document: default
  html_document: default
---

AUTHOR LIST:  

Josh G. Chenoweth^1qc^, Carlo Colantuoni^2q^, Deborah A. Striegel^1^, Pavol Genzor^1^, Joost Brandsma^1^, Paul W. Blair^1,3^, Subramaniam Krishnan^1^, Elizabeth Chiyka^1^, Mehran Fazli^1^, Rittal Mehta^1^, Michael Considine^4^, Leslie Cope^4^, Audrey C. Knight^2^, Anissa Elayadi^1^, Anne Fox^5^, Ronna Hertzano^6,7^, Andrew G. Letizia^5^, Alex Owusu-Ofori^8,9^, Isaac Boakye^10^, Albert A. Aduboffour^8^, Daniel Ansong^11,12^, Eno Biney^13^, George Oduro^13^, Kevin L. Schully^14^, Danielle V. Clark^1^.

####

^1^Austere environments Consortium for Enhanced Sepsis Outcomes (ACESO), The Henry M. Jackson Foundation for the Advancement of Military Medicine, Inc., Bethesda, Maryland, USA.  
^2^Department of Neurology, Johns Hopkins School of Medicine, Baltimore, MD, USA.  
^3^Department of Pathology, Uniformed Services University, Bethesda, MD.  
^4^Sidney Kimmel Comprehensive Cancer Center, Johns Hopkins University, Baltimore, MD.  
^5^Naval Medical Research Unit EURAFCENT Ghana detachment, Accra, Ghana.  
^6^Institute for Genome Sciences, University of Maryland School of Medicine, Baltimore, MD, USA.  
^7^Departments of Otorhinolaryngology Head and Neck Surgery, Anatomy and Neurobiology, University of Maryland School of Medicine, Baltimore, MD, USA.  
^8^Laboratory Services Directorate, Komfo Anokye Teaching Hospital (KATH), Kumasi, Ghana.  
^9^Department of Clinical Microbiology, Kwame Nkrumah University of Science and Technology (KNUST), Ghana.  
^10^Research and Development Unit, KATH, Kumasi, Ghana.  
^11^Child Health Directorate, KATH, Kumasi, Ghana.  
^12^Department of Child Health, KNUST, Kumasi, Ghana.  
^13^Accident and Emergency Department, KATH, Kumasi, Ghana.  
^14^Austere environments Consortium for Enhanced Sepsis Outcomes (ACESO), Biological Defense Research Directorate, Naval Medical Research Command-Frederick, Ft. Detrick, Maryland, USA.  
  
  
__These authors contributed equally:__ Josh G. Chenoweth and Carlo Colantuoni
  
####

Compiled by: __Pavol Genzor__  
Date: __January 22nd, 2024__  
Version: V2

##
### Summary 

This document contains an example of the code used to analyze data associated with the manuscript. The "transfer learing" procedure consist of three main steps specific to each use where in general:  

* 1. Expression data is subjected to negative matrix factorization using __CoGAPS__
[**[link]**](https://bmcbioinformatics.biomedcentral.com/articles/10.1186/s12859-020-03796-9). 

* 2. Public data of interest is acquired and prepared (*NOTE: To obtain public data please see original study or contact the corresponding authors.*)

    + Sepsis:
      + Reyes et.al., Nature Medicine, 2020 
      + Tsalik et.al., Science Translational Medicine, 2016
      + Cazalis et.al., Intensive Care Medicine Experimental, 2014  
      
    + COVID-19:
      + Reyes et.al., Science Translational Medicine, 2021

    
* 3. Decomposed data is projected onto a public data using __projectR__
[**[link]**](https://www.bioconductor.org/packages/release/bioc/html/projectR.html) 

* 4. Results are plotted using preferred software.


##
### Procedure details 


##
### Step1:  
Install the [___***CoGAPS***___ package](https://www.bioconductor.org/packages/release/bioc/html/CoGAPS.html) and load it into the environment. Next, load your normalized log2 transformed data into R environment (**DATi**). Make sure that data is configured in a matrix object where _rownames_ correspond to genes names and _colnames_ correspond to cells/samples. Use this data to generate _prcomp_ object that will later be used for decomposition. Finally, load the data table will all relevant metadata information. Once data is loaded, configure the running parameters for the CoGAPS (see below exact parameters used in manuscript) and run the algorithm and perform data decomposition.  

##
```{r, load & CoGAPS, eval=FALSE}
## Install and load required package

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("CoGAPS")

# Load the library

library("CoGAPS")

## Load the expression data

load("path/to/your/directory/your_expression_data_and_metadata.rda")

## Input should be matrix of log2 transformed data where:
## - genes as row names
## - cell/sample names as column names

DATi
cells <- colnames(DATi)
genes <- rownames(DATi)

## Table containing experiment-related metadata that will be used for plotting

sampORIG

## Run principal component analysis on the data set
## - PCA can be used in the projection in step 3

pca <- prcomp(t(DATi))
pca

## Set the analysis parameters

nThrd=1
params <- new("CogapsParams")
params <- setParam(params, "sparseOptimization", FALSE)
params <- setParam(params, "nIterations", 40000)
params <- setParam(params, "nPatterns", 30)
params <- setDistributedParams(params, nSets=1)

## Run the analysis using following parameters 

xxCoGAPS <- CoGAPS(data=DATi,
                   params=params,
                   geneNames=genes,
                   sampleNames=cells,
                   messages=TRUE,
                   transposeData=FALSE,
                   nPatterns=30,
                   nIterations=40000,
                   nThreads=nThrd)

## NOTE: missing objects
#pca
#pcvars
#sampORIG

```

## 
### Step2:  
Attain the public data set of interest making sure it contains all information of interest. You may need to contact authors if the information is incomplete. Format the data for use with the [___***projectR***___](https://www.bioconductor.org/packages/release/bioc/html/projectR.html). This may involve synchronizing the nomenclature, ensuring data is normalized, and most importatnly, organized into matrix with appropriate _rownames_ and _colnames_. 

+ Data used in this manuscript:
  + Reyes et.al., Science Translational Medicine, 2021 - COVID-19 - [**[link]**](https://www.science.org/doi/10.1126/scitranslmed.abe9599)
  + Reyes et.al., Nature Medicine, 2020 - Sepsis - [**[link]**](https://www.nature.com/articles/s41591-020-0752-4)
  + Tsalik et.al., Science Translational Medicine, 2016 - Bacterial vs Viral - [**[link]**](https://www.science.org/doi/abs/10.1126/scitranslmed.aad6873)
  + Cazalis et.al., Intensive Care Medicine Experimental, 2014, - Sepsis shock - [**[link]**](https://icm-experimental.springeropen.com/articles/10.1186/s40635-014-0020-3)
  


For each data set we used we generated following objects: __seq__ - a _matrix_ object of relevant expression data that has been log2 transformed; __samples__ - a _vector_ containing cell IDs or sample names depending on experiment; __genes__ - a vector object containing gene/feature names. 

##
```{r prepare public data, eval=FALSE}
## Load prepared public data
## - preparation entails obtaining the expression information and metadata
## - required components include the gene information, and cell/sample information
## - processed public data can be stored in R object for future use

load("BYOD/public_data_A.RData")

## Content

seq       #: expression matrix
samples   #: cell/sample -> column names
genes     #: genes -> rownames

## To coerce data.frame to matrix
## - "seq" object must be matrix NOT a data.frame

if(class(seq)=="data.frame"){
  seq=as.matrix(seq)}

## To log2 scale the raw data
## - add 1 and apply log2 normalization

if(range(seq)[2]>100){
  print("max<100 - log2'ing")
  seq=log2(seq+1)}

```

##
### Step3:  
Project your data onto the public data set using [___***projectR***___](https://www.bioconductor.org/packages/release/bioc/html/projectR.html). To get more information about configurations please reefer to the software vignette. [HERE](https://www.bioconductor.org/packages/release/bioc/vignettes/projectR/inst/doc/projectR.pdf). Our sample configurations are shown below. 

##
```{r, protejct project, eval=FALSE}
## Install an load required package

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("projectR")

# Load the library

library(projectR)

## Extract A values from CoGAPS object
## - this was prepared in Step1 above

As = xxCoGAPS$featureLoadings


## Project first two principal components
## - "seq" corresponds to the public data set
## - "loadings" correspond to pca 

PC.xproj=projectR(data=seq,
                  loadings=pca,
                  NP=1:2)

## PLOT the projections
## - can be accomplished via any method (baseR or ggplot graphics)

```

##
### NOTES
##

```{r, sample run, eval=TRUE}
## NOTE:
# -write notes here
```











